Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-07-30 ms/manuscript.Rnw
Some of the changes suggested by @dfalster

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low carbon budget and death independent of carbon budget -- and how a species' light demand, maximum dbh and wood density affect each. We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species. We show that carbon-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island. Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of carbon-related hazards and carbon-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into carbon-dependent and carbon-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations. Moreover, we show that while these three traits do affect both carbon-dependent and carbon-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}. The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire. An individual's total hazard at a given time point is therefore the sum of these two independent hazards. However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamanetal hazards faced by plants. Instead, plant mortality is typically modelled using: 1) logistic regression (i.e. with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.

@@

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
B) We consider three alternative hazard functions: 1) a carbon-independent baseline hazard (i.e. average hazard), 2) a carbon-dependent hazard, where growth rate is a proxy for carbon budget; and 3) a function that combines both carbon-dependent and independent hazards. The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e. low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right). In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture carbon-dependent hazards, while $\gamma$ captures carbon-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g. windfall, fire). For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for carbon-dependent and carbon-independent hazards. We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Carbon independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light rdemand ($\upsilon$) and maximum dbh ($\psi$). We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
A) Six sequential stages of model selection with each stage increasing in complexity. \textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e. carbon budget). This includes two possible hazard functions: a carbon-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + carbon-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters. Symbols represent the three hazard functional form: baseline (circle), carbon-dependent (triangle), and hazard containing both carbon-independent and carbon-dependent hazards (square).}
\label{fig2}
\end{figure*}

To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e. a null model) and 2) a carbon-dependent only hazard. We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across six iterative stages of model development.

@@

In stage 2, we assessed the predictive performance of three alternative hazard functions (Fig. \ref{fig1}B). The simplest form (Fig. \ref{fig1}B, left) assumes a constant hazard. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases. The third form (Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some carbon-independent baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared. In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts carbon-dependent mortality.

In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance. Specifically, we examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), as well as 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters. Lastly, we fit a model that included a species random effect on each parameter (stage 6). This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits. Because these models were computationally expensive to run (with some models requiring over 400 computing hours), we were unable to examine interactions among traits or determine the most predictive combination of traits effects and parameters in a cross validated framework.

@@

\subsection*{Mortality over time}
Examining the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time. The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding the term $\delta_t$) we observed a small increase in predictive skill (Fig. \ref{fig2}A). Table S1 gives individual census effects.

\subsection*{Hazard functions}
Comparing the three hazard functions summarised in Fig. \ref{fig1}B, we found that the third function --- the function containing both carbon-dependent and carbon-independent hazard terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and carbon-dependent only hazard functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig. \ref{fig2}A). Table S1 gives a summary of mean hyper parameter estimates.

\begin{figure*}[ht!]
\includegraphics{figures/fig3.pdf}
\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$ (low growth effect), $\beta$ (growth-mortality decay rate) and $\gamma$ (baseline/carbon-independent hazard). Points are the estimated mean ($\pm$ 95\% credible intervals) parameters for each of the 203 Barro Colorado Island species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship across observed trait ranges.}
\label{fig3}
\end{figure*}

\subsection*{Species traits}
Using the hazard function form with highest predictive skill (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$; Fig.~\ref{fig1}b right), we examined how effects of a species' maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) placed on $\alpha$, $\beta$ and $\gamma$ affected model predictive accuracy. In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}). Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits. However, this increase in predictive accuracy was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6). See Figs. S3-S5 for species-level parameter estimates.

@@

In order to examine how species traits -- maximum dbh, light demand and wood density -- affect tree mortality rates, we fitted a final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$. We found that each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Specifically, we found that light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower carbon-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower carbon-independent hazards. Table S1 for trait effects on each parameter.
\subsection*{Parameter relative importance}

@@

We then examined the relative contribution species traits, random effects and carbon-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC. First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e. the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the carbon-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the carbon-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the carbon-dependent hazard resulted in substantially lower predictive ability.

\section*{Discussion}
Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into carbon-dependent and carbon-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability. Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.

The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both carbon-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill. This is because the carbon-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g. competition, parasites, herbivory). By contrast, the carbon-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g. windthrow, lightning strike).

Additionally, the partitioning of mortality rates into both carbon-dependent and carbon-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each. Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that that the removal of the carbon-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}). This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.

Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{Poorter:2006tq, Chave:2009iy}. Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.

@@

@@

\showacknow % Display the acknowledgements section

\vspace{-5em}
\pnasbreak

% Bibliography
\bibliography{refs.bib}
%\pnasbreak
\end{document}

>>>>

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each. We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species. We show that growth-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island. Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related hazards and growth-independent hazards on the overall mortality rates. We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations. Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species. Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}. The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair. The second are growth-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire. An individual's total hazard at a given time point is therefore the sum of these two independent hazards. However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamanetal hazards faced by plants. Instead, plant mortality is typically modelled using: 1) logistic regression (i.e. with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.

@@

\begin{figure}[!hb]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology.
A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e. average hazard; Panel A; left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (Panel A; middle); and 3) a function that combines both growth-dependent and independent hazards (Panel A; right). The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e. low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote. In the most complex model (Panel A; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g. windfall, fire). For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
D) The predictive skill of alternative models was evaluated via 10-fold cross validation. The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$). Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data. Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
\label{fig1}
\end{figure}

Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for growth-dependent and growth-independent hazards. We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for the additive, thus independent, effects of growth-dependent and growth-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$. Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$. Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light rdemand ($\upsilon$) and maximum dbh ($\psi$). We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
A) Six sequential stages of model selection with each stage increasing in complexity. \textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e. carbon budget). This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters. Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure*}

To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e. a null model) and 2) a growth-dependent only hazard. We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken approximately at 5-year intervals over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across six iterative stages of model development.

@@

In stage 2, we assessed the predictive performance of three alternative hazard functions (Fig. \ref{fig1}B). The simplest form (Fig. \ref{fig1}B, left) assumes a constant hazard. The second form (Fig. \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases. The third form (Fig. \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some growth-independent baseline hazard $\textgreater 0$. Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared. In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.

In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how three species traits considered major explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}: wood density, light demand and maximum dbh, affected model predictive performance. Specifically, we examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), as well as 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters. Lastly, we fit a model that included a species random effect on each parameter (stage 6). This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits. Because these models were computationally expensive to run (with some models requiring over 400 computing hours), we were unable to examine interactions among traits or determine the most predictive combination of traits effects and parameters in a cross validated framework.

@@

\subsection*{Mortality over time}
Examining the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time. The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding the term $\delta_t$) we observed a small increase in predictive skill (Fig. \ref{fig2}A). Table S1 provides individual census effects.

\subsection*{Hazard functions}
Comparing the three hazard functions summarised in Fig. \ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig. \ref{fig2}). Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig. \ref{fig2}A). Table S1 gives a summary of mean parameter estimates.

\begin{figure*}[ht!]
\includegraphics{figures/fig3.pdf}
\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$ (low growth effect), $\beta$ (growth-mortality decay rate) and $\gamma$ (baseline/growth-independent hazard). Points are the estimated mean ($\pm$ 95\% credible intervals) parameters for each of the 203 Barro Colorado Island species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship across observed trait ranges.}
\label{fig3}
\end{figure*}

\subsection*{Species traits}
Using the hazard function form with highest predictive skill (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$; Fig.~\ref{fig1}b right), we examined how effects of a species' maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) placed on $\alpha$, $\beta$ and $\gamma$ affected model predictive accuracy. In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}). Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits. However, this increase in predictive accuracy was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6). See Tables S1 for trait coefficients and Table S2 or Figs. S3-S5 for species-level parameter estimates.

@@

In order to examine how species traits -- maximum dbh, light demand and wood density -- affect tree mortality rates, we fitted a final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$. We found that each trait influenced tree mortality in different ways (Fig. \ref{fig3}-\ref{fig4}). Specifically, we found that light demanding species were more susceptible to low growth rates (i.e. higher $\alpha$), had higher growth-independent hazards (i.e. higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e. lower $\beta$). By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards. Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards. Table S1 for trait effects on each parameter.
\subsection*{Parameter relative importance}

@@

We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC. First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e. the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.

\section*{Discussion}
Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability. Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.

The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}. We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill. This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g. competition, parasites, herbivory). By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g. windthrow, lightning strike).

Additionally, the partitioning of mortality rates into both growth-dependent and growth-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each. Like many other studies \cite{King:1994vg,king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates. Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}). This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.

Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}. Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.

@@

Weak correlation (\textless 0.36) was found between species' wood density, light demand and maximum dbh (See Fig. S6)

@@

\showacknow % Display the acknowledgements section
\vspace{-5em}

\newpage
% Bibliography
\bibliography{refs.bib}

\end{document}
